An efficient feedback calibration algorithm for direct imaging radio telescopes

### 2016-02-22 cal_demo/ms.tex
Add comments to section 2
(*) Comments are in bold face

The MOFF algorithm next calls for a calibration. We will assume we have formed an estimate of the gains after $n$ iterations of a calibration loop, and derive an updated $n+1$ estimate. How the calibration gain is applied can depend on the dominant source of the noise in the measurement. If the noise is dominated by sky fluctuations, the noise level is affected by the antenna gains so dividing by the complex gains will flatten the noise across antennas. However, if the noise is dominated in the receiver (after the complex gain factor of equation~\ref{eq:apply_gain}), a lower amplitude gain results in lower signal to noise for that antenna. Therefore, one should multiply by the conjugate of the gain to down weight appropriately while removing the phase acquired. We will abstain from this choice for now, and instead use a multiplicative factor $h^{(n)}_a$ to represent the application of our $n^{\mathrm{th}}$-loop estimate of the gain for antenna $a$:
\begin{equation}
\Erest{a} = h^{(n)}_a \Er{a}
\end{equation}
where,
\begin{equation}
h^{(n)}_a=\begin{cases}
1/g^{(n)}_a & \mbox{sky noise dominated} \\ 
g^{*(n)}_a & \mbox{receiver noise dominated}.
\end{cases}
\end{equation}

A dirty image is next formed by gridding the calibrated fields with the antenna voltage pattern, Fourier transforming, squaring, and averaging in time. The estimated value for a pixel, $\s_i$, can be expressed as
\begin{equation}
I'(\s_i) = \left<\left| \frac{1}{\Nant}\sum_i e^{2\pi i \mathbf{r}\cdot \s_i} \sum_a \beamr_a(\ri - \ra) h^{(n)}_a g^T_a \Er{a}^T(t) \right|^2 \right>_t.
\end{equation}
This is the final output of the MOFF correlator. However, for calibration purposes we will be interested in the electric field image just prior to squaring and averaging.
\begin{equation}
\Ethetaest (\s_i,t) = \frac{1}{\Nant} \sum_i e^{2\pi i \ri \cdot \s_i} \sum_a \beamr_a(\ri - \ra) h^{(n)}_a g^T_a \Er{a}^T(t)
\end{equation}
We can simplify this expression by exchanging the sums to transform the beam term into sky coordinates.
\begin{align}\label{eq:epix}
\Ethetaest(\s_i,t) & = \frac{1}{\Nant} \sum_a h^{(n)}_a g^T_a\Er{a}^T(t) e^{2\pi i \s_i \cdot \ra}\sum_i \beamr_b(\ri-\ra)e^{2\pi i \s_i \cdot (\ri-\ra)} \nonumber\\
& = \frac{1}{\Nant} \sum_a h^{(n)}_a g^T_a\Er{a}^T(t) e^{2\pi i \s_i \cdot \ra}\beamtheta_a(\s_i)
\end{align}

@@

Plugging equation \ref{eq:epix} into equation \ref{eq:Cna}, we find,
\begin{align}\label{eq:cna}
\Cna & = \left<g^T_a \Er{a}^T(t) \frac{1}{\Nant} \sum_b h^{*(n)}_b g^{T*}_b\Er{b}^{T*}(t) e^{-2\pi i \spix \cdot \rb} \beamtheta^*_b(\spix)\right>_t \nonumber \\
& = \frac{g^T_a}{\Nant} \sum_b h^{*(n)}_b g^{T*}_b \beamtheta^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \left<\Er{a}^T \Er{b}^{*T} \right>_t \nonumber \\
& = \frac{g^T_a}{\Nant} \sum_b h^{*(n)}_b g^{T*}_b \beamtheta^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \V^T_{ab}
\end{align}
where in the second step we group time-dependent terms, and in the third we define the true visibilities as the correlation between true antenna electric field measurements. It is easy to see from here that the net effect of including the receiver noise term, $\widetilde{n}_a(f,t)$, will result in added noise on the true visibilities, including a bias on the auto-correlation terms. In principle a bias can result on any visibility from any noise correlations between antennas, but the implementation included in the EPIC software package restricts the noise term to include only the auto-correlation noise bias, and assumes baseline dependent noise is zero mean.

@@

While testing we found equation~\ref{eq:cal_solution} resulted in oscillatory gain solutions as it was iterated, as is often the case in iterative minimization methods. To mitigate this we introduce a damping factor, $0 \leq \damp <1$, which is used to attenuate the gain update, effectively giving the solutions memory of previous iterations.
\begin{equation}
g^{(n+1)}_a = (1-\damp) g'^{(n+1)}_a + \damp g^{(n)}_a
\end{equation}
We found that while equation~\ref{eq:cal_solution} does indeed converge on good solutions, the process is made faster by tuning the damping factor. While $g'^{(n+1)}_a$ is the best estimate of the gains after the $n^{\textrm{th}}$ iteration, the damped version, $g^{(n+1)}_a$, is actually used in the iterative calibration loop. Once the loop converges the difference is a longer effective integration for the damped version (lower thermal noise).

We show schematically the process of calibrating a direct imaging correlator in figure~\ref{fig:schematic}. Computationally expensive steps that must be performed ``on-chip" are shown inside the gray box. The uncalibrated antenna signals are tapped out after the F-engine and correlated against the output image pixel of interest. The correlated values are then passed off-chip to estimate the gains using equation~\ref{eq:cal_solution}, and additional fitting if desired. The gains are then passed back to the correlator to update the calibration for subsequent data streams.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figures/schematic.pdf}
\caption{The general data flow of the MOFF correlator, with a feedback calibration loop. A pixel from the (unsquared) image is tapped out and correlated against the input antenna electric field signals to form $\Cna$ coefficients, which are used to estimate the antenna gains. These gain estimates are then fed back into the correlator to be applied to subsequent data, and the process is repeated. The gray box shows operations which must be done at high speed (before averaging in time), which white boxes show operations which can be performed ``off-chip".}
\label{fig:schematic}
\end{center}
\end{figure}

An important feature to note is that, like the MOFF-generated images themselves, equations~\ref{eq:cal_solution} and~\ref{eq:cal_solution_simple} include the antenna auto-correlations (the sum is over \emph{all} $b$, not excluding $a$). It can be difficult to perfectly model the noise bias from auto-correlations, which can often times be far brighter than the visibilities themselves. It can therefore be beneficial to subtract this term directly from \Cna, and exclude the $b=a$ term in the sum.
\begin{equation}
\Cna \rightarrow \Cna - \frac{1}{\Nant} h^{*(n)}_a\beamtheta^*_a e^{-2\pi i \spix \cdot \ra} \left<|\Er{a}|^2\right>_t
\end{equation}
This requires generating these correlations, which again only scale as $\mathcal{O}(\Nant)$, and are generally useful for array diagnostics.

We conclude this section by connecting our calibration expression to that found in a visibility framework. In the limit of a single bright calibrating source at phase center, we can greatly simplify equations~\ref{eq:cna} and~\ref{eq:cal_solution}. We will assume the beams are normalized such that $\beamtheta(0)=1$. We can further drop the exponential phase terms because $\spix=0$. We then absorb the true gains and gain corrections into the true visibilities in equation~\ref{eq:cna} to express as a sum of measured visibilities.
\begin{equation}
\mathcal{C}^{(n)}_{a,0} \rightarrow \frac{1}{\Nant}\sum_b h^{*(n)}_b \V_{ab}
\end{equation}

>>>>

The MOFF algorithm next calls for a calibration. We will assume we have formed an estimate of the gains after $n$ iterations of a calibration loop, and derive an updated {\bf estimate for the ($n+1$)-{\it th} iteration}. How the calibration gain is applied can depend on the dominant source of the noise in the measurement. If the noise is dominated by sky fluctuations, the noise level is affected by the antenna gains so dividing by the complex gains will flatten the noise across antennas. However, if the noise is dominated in the receiver (after the complex gain factor of equation~\ref{eq:apply_gain}), a lower amplitude gain results in lower signal to noise for that antenna. Therefore, one should multiply by the conjugate of the gain to down weight {\bf it} appropriately while removing the phase acquired. We will abstain from this choice for now, and instead use a multiplicative factor $h^{(n)}_a$ to represent the application of our $n^{\mathrm{th}}$-loop estimate of the gain for antenna $a$:
\begin{equation}
\Erest{a} = h^{(n)}_a \Er{a}
\end{equation}
where,
\begin{equation}
h^{(n)}_a=\begin{cases}
1/g^{(n)}_a, & \mbox{sky noise dominated} \\ 
g^{*(n)}_a, & \mbox{receiver noise dominated}.
\end{cases}
\end{equation}

A dirty image is next formed by gridding the calibrated fields with the antenna voltage pattern, Fourier transforming, squaring, and averaging in time. The estimated value for a pixel, $\s_i$, can be expressed as
\begin{equation}
I'(\s_i) = \left<\left| \frac{1}{\Nant}\sum_i e^{2\pi i \mathbf{r}\cdot \s_i} \sum_a \beamr_a(\ri - \ra) h^{(n)}_a g^T_a \Er{a}^T(t) \right|^2 \right>_t.
\end{equation}
This is the final output of the MOFF correlator. However, for calibration purposes we will be interested in the electric field image just prior to squaring and averaging.
\begin{equation}
\Ethetaest (\s_i,t) = \frac{1}{\Nant} \sum_i e^{2\pi i \ri \cdot \s_i} \sum_a \beamr_a(\ri - \ra) h^{(n)}_a g^T_a \Er{a}^T(t)
\end{equation}
We can simplify this expression by exchanging the sums to transform the beam term into sky coordinates.
\begin{align}\label{eq:epix}
\Ethetaest(\s_i,t) & = \frac{1}{\Nant} \sum_a h^{(n)}_a g^T_a\Er{a}^T(t) e^{2\pi i \s_i \cdot \ra} \nonumber \\ 
  & \qquad\qquad \sum_i \beamr_b(\ri-\ra)e^{2\pi i \s_i \cdot (\ri-\ra)} \nonumber\\
& = \frac{1}{\Nant} \sum_a h^{(n)}_a g^T_a\Er{a}^T(t) e^{2\pi i \s_i \cdot \ra}\beamtheta_a(\s_i)
\end{align}

@@

Plugging equation \ref{eq:epix} into equation \ref{eq:Cna}, we find,
\begin{align}\label{eq:cna}
\Cna & = \left<g^T_a \Er{a}^T(t) \frac{1}{\Nant} \sum_b h^{*(n)}_b g^{T*}_b\Er{b}^{T*}(t) e^{-2\pi i \spix \cdot \rb} \beamtheta^*_b(\spix)\right>_t \nonumber \\
& = \frac{g^T_a}{\Nant} \sum_b h^{*(n)}_b g^{T*}_b \beamtheta^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \left<\Er{a}^T \Er{b}^{*T} \right>_t \nonumber \\
& = \frac{g^T_a}{\Nant} \sum_b h^{*(n)}_b g^{T*}_b \beamtheta^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \V^T_{ab}
\end{align}
where in the second step we group time-dependent terms, and in the third we define the true visibilities as the correlation between true antenna electric field measurements. It is easy to see from here that the net effect of including the receiver noise term, $\widetilde{n}_a(f,t)$, will result in added noise on the true visibilities, including a bias on the auto-correlation terms. In principle a bias can result on any visibility from any noise correlations between antennas, but the implementation included in the EPIC software package restricts the noise term to include only the auto-correlation noise bias, and assumes baseline dependent noise {\bf has} zero mean.

@@

While testing we found equation~\ref{eq:cal_solution} resulted in oscillatory gain solutions as it was iterated, as is often the case in iterative minimization methods. To mitigate this we introduce a damping factor, $0 \leq \damp <1$, which is used to attenuate the gain update, effectively giving the solutions memory of previous iterations.
\begin{equation}
g^{(n+1)}_a = (1-\damp) g'^{(n+1)}_a + \damp g^{(n)}_a
\end{equation}
We found that while equation~\ref{eq:cal_solution} does indeed converge on good solutions, the process is made faster by tuning the damping factor. While $g'^{(n+1)}_a$ is the best estimate of the gains after the $n^{\textrm{th}}$ iteration, the damped version, $g^{(n+1)}_a$, is actually used in the iterative calibration loop. Once the loop converges the difference is a longer effective integration for the damped version (lower thermal noise) ({\bf the last sentence is too condensed that it is not clear what you are trying to say}).

We show schematically the process of calibrating a direct imaging correlator in figure~\ref{fig:schematic}. Computationally expensive steps that must be performed ``on-chip" are shown inside the gray box. The uncalibrated antenna signals are tapped out after the F-engine and correlated against the output image pixel of interest ({\bf i.e., equation~\ref{eq:Cna} is ``on-chip''}). The correlated values are then passed off-chip to estimate the gains using equation~\ref{eq:cal_solution}, and additional fitting if desired. The gains are then passed back to the correlator to update the calibration for subsequent data streams.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figures/schematic.pdf}
\caption{The general data flow of the MOFF correlator, with a feedback calibration loop. A pixel from the (unsquared) image is tapped out and correlated against the input antenna electric field signals to form $\Cna$ coefficients, which are used to estimate the antenna gains. These gain estimates are then fed back into the correlator to be applied to subsequent data, and the process is repeated. The gray box shows operations which must be done at high speed (before averaging in time), which white boxes show operations which can be performed ``off-chip". ({\bf I have some suggestions to tweak this figure.})}
\label{fig:schematic}
\end{center}
\end{figure}

An important feature to note is that, like the MOFF-generated images themselves, equations~\ref{eq:cal_solution} and~\ref{eq:cal_solution_simple} include the antenna auto-correlations (the sum is over \emph{all} $b$, not excluding $a$). It can be difficult to perfectly model the noise bias from auto-correlations, which can often times be far brighter than the visibilities themselves. It can therefore be beneficial to subtract this term directly from \Cna, and exclude the $b=a$ term in the sum.
\begin{equation}
\Cna \rightarrow \Cna - \frac{1}{\Nant} h^{*(n)}_a\beamtheta^*_a e^{-2\pi i \spix \cdot \ra} \left<|\Er{a}|^2\right>_t
\end{equation}
This requires generating these correlations, which again only scale as $\mathcal{O}(\Nant)$, and are generally useful for array diagnostics. ({\bf What happened to the $g_a^T$ term outside the sum? And why did you drop $\hat{\mathbf{s}}_0$ in $W_a^*$}?)

We conclude this section by connecting our calibration expression to that found in a visibility framework. In the limit of a single bright calibrating source at phase center, we can greatly simplify equations~\ref{eq:cna} and~\ref{eq:cal_solution}. We will assume the beams are normalized such that $\beamtheta(0)=1$. We can further drop the exponential phase terms because $\spix=0$. We then absorb the true gains and gain corrections into the true visibilities in equation~\ref{eq:cna} to express as a sum of measured visibilities.
\begin{equation}
\mathcal{C}^{(n)}_{a,0} \rightarrow \frac{1}{\Nant}\sum_b h^{*(n)}_b \V_{ab}
\end{equation}
({\bf It is not clear what happened to the $g$ terms.})
